### 2016-04-26 Paper_v5.tex
Updates to the introduction

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Dynamic Optimization with Particle Swarms (DOPS): A meta-heuristic for parameter estimation in biochemical models}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker$^{\dag}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {$^{\dag}$School of Civil and Environmental Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{PLoS ONE}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Mathematical modeling is a powerful tool to analyze, and ultimately design biochemical networks.
However, the estimation of biochemical model parameters is a significant challenge.
Parameter estimation in biochemical models typically involves expensive function evaluations and noisy data, making it difficult to quickly obtain optimal solutions.
Biochemical models often also have many local extrema which further complicates parameter estimation.
Toward these challenges, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel global meta-heuristic that combined features of multi-swarm particle swarm optimization with dynamically dimensioned search (DDS).
DOPS uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors, the best of which is greedily updated using dynamically dimensioned search.
We tested the performance of DOPS on a model of human coagulation cascade.
We performed $\mathcal{T}$ = 25 trials with $\mathcal{N}$ = 4000 function evaluations per trial, and compared the performance of DOPS with other commonly
used meta-heuristics such as differential evolution (DE), simulated annealing (SA) and dynamically dimensioned search (DDS).
We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between simulations and experimental measurements.
Lastly, we tested the performance of DOPS on commonly used test functions for global optimization and on published biochemical parameter estimation benchmark problems.
For the wide range of problems that we considered, DOPS outperformed other meta-heuristic approaches despite a limited number of function evaluations.
Taken together, DOPS is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.

@@

Cells process nutrients and respond to changes in their environment using complex biochemical networks. These networks contain thousands of components
interconnected through nonlinear enzyme catalyzed reactions. Mathematical modeling has evolved as a powerful paradigm to analyze, and ultimately design these complex networks \cite{assmus2006dynamics, Riel2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of biochemical networks is often an iterative process.
First, models are formulated from biochemical knowledge, and then model parameters are estimated using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
Parameter estimation is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and model simulations is minimized using an optimization strategy \cite{moles2003parameter}. Optimal parameters obtained from model training are then used to validate the model on unseen experimental data.
If validation fails, model construction and calibration are repeated iteratively until satisfactory results are obtained.

Parameter estimation is a major challenge in the iterative development of biochemical models.
Although parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a unique estimation challenge.
Often optimization problems involving biological systems are non-linear and multi-modal i.e. typical models often have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}.
Non-linearity coupled with multi-modality generally renders local optimization techniques such as pattern search \cite{hooke1961direct}, Nelder-Mead simplex methods \cite{nelder1965simplex},
steepest descent methods or Levenberg-Marquardt \cite{more1978levenberg} incapable of reliably obtaining optimal solutions as they generally terminate at local minimum.
Though deterministic global optimization techniques (for example algorithms based on branch and bound) can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global},
the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of knowledge about the objective function hampers these techniques.

Meta-heuristic stochastic optimization approaches like Genetic Algorithms (GAs) \cite{goldberg2006genetic}, Simulated Annealing (SA) \cite{kirkpatrick1983optimization},
Evolutionary Programming \cite{fogel2009artificial} and population based searches like Differential Evolution (DE) \cite{storn1997differential} have shown promise on nonlinear multi-modal problems \cite{sun2012parameter}.
These techniques do not make any assumptions about the structure of objective function nor do they require \textit{a~priori} information about the objective function.
Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.
Mendes et al. \cite{mendes1998non} used simulated annealing to estimate rate constants for the irreversible inhibition of HIV proteinase,
Modchang et al. \cite{modchang2008mathematical} used a genetic algorithm (GA) to estimate parameters for a model of G-protein-coupled receptors (GPCRs) mediated signal transduction. The parameters obtained using GA helped qualitatively ascertain the efficacy and potency of two G-protein agonists - N6-cyclopentyladenosine (CPA) and 5'-N-ethylcarboxamidoadenosine (NECA). 
Differential evolution approaches have also been effective on various biological problems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}.
Tashkova et al. compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective \cite{tashkova2011parameter}.
Banga and co-workers have also successfully applied scatter-search methods to estimate parameters on non-linear biological processes \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter}.
Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near globally optimal solution obtained using a meta-heuristic is refined using a local search have also become popular.
Villaverde et al. combined scatter search with local search methods for parameter estimation in a collection of systems biology models \cite{villaverde2015biopredyn}.
Fan et al. recently showed that population based meta-heuristics along with decomposition based methods can be also used to model gene circuits from mRNA data \cite{fan2015parameter}.
Despite these successes, a major drawback of most metaheuristic approaches is the large number of function evaluations required to explore the parameter space.
Typically as models grow in size and complexity, evaluation of the objective function becomes computationally expensive.
Thus performing a large number of function evaluations is not desirable (and perhaps not feasible).

In many of these high dimensional problems approaching an exact solution may not be necessary. Gutenkust et al. \cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'. Sloppy systems have specific parameter combinations that largely define the dynamics of the system. Large perturbations to the rest of the parameters does not greatly impact the system dynamics. Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}. Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions. DDS is a single solution based search where the solution is greedily updated. Because the neighborhood in DDS is the entire parameter space, DDS is a global optimization technique although it is greedy.

In this study, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel meta-heuristic that combines the global search capability of multi-swarm particle swarm optimization and dynamically dimensioned search (DDS). The objective of DOPS is to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
DOPS uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamically dimensioned search. We first considered a model of human coagulation cascade to test the performance of DOPS. Coagulation is a large, complex biochemical network involving strong positive feedback.
We then tested the performance of DOPS on commonly used test functions for global optimization (Ackley and Rosenbrock) and published biochemical parameter estimation benchmark problems \cite{villaverde2015biopredyn}. DOPS outperformed common meta-heuristic approaches like Differential Evolution (DE), Simulated Annealing (SA) and dynamically dimensioned search (DDS) on both the test functions and the coagulation model.
We also compared the performance of DOPS against enhanced scatter search (eSS) on two benchmark problems published by Villaverde and co-workers\cite{villaverde2015biopredyn}. We fixed the number of function evaluations to 4000 for both the benchmark problems without pre-optimizing any algorithmic parameters. DOPS recovered the nominal parameters and surprisingly did so faster than eSS which took nearly an order of function evaluations more than DOPS in both the cases.
Taken together, these studies suggest DOPS is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Dynamic Optimization with Particle Swarms (DOPS): A meta-heuristic for parameter estimation in biochemical models}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker$^{\dag}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {$^{\dag}$School of Civil and Environmental Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Biotechnology~Journal}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Mathematical modeling is a powerful tool to analyze, and ultimately design biochemical networks.
However, the estimation of biochemical model parameters is a significant challenge.
Parameter estimation in biochemical models typically involves expensive function evaluations and noisy data, making it difficult to quickly obtain optimal solutions.
Biochemical models often also have many local extrema which further complicates parameter estimation.
Toward these challenges, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel global meta-heuristic that combined features of multi-swarm particle swarm optimization with dynamically dimensioned search (DDS).
DOPS uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors, the best of which is greedily updated using dynamically dimensioned search.
We tested the performance of DOPS on a model of human coagulation cascade.
We performed $\mathcal{T}$ = 25 trials with $\mathcal{N}$ = 4000 function evaluations per trial, and compared the performance of DOPS with other commonly
used meta-heuristics such as differential evolution (DE), simulated annealing (SA) and dynamically dimensioned search (DDS).
We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between simulations and experimental measurements. Lastly, we tested the performance of DOPS on commonly used test functions for global optimization and on published biochemical parameter estimation benchmark problems. For the wide range of problems that we considered, DOPS outperformed other meta-heuristic approaches despite a limited number of function evaluations. Taken together, DOPS is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.

@@

Cells process nutrients and respond to changes in their environment using complex enzyme catalyzed biochemical networks.
Mathematical modeling has evolved as a powerful paradigm to analyze, and ultimately design these complex networks \cite{assmus2006dynamics, Riel2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of biochemical networks is often an iterative process.
First, models are formulated from existing biochemical knowledge, and then model parameters are estimated using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
Parameter estimation is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental measurements and model simulations is minimized using an optimization strategy \cite{moles2003parameter}.
Optimal parameter estimates obtained from model training are then used to validate the model on unseen experimental data.
If the validation studies fail, model construction and calibration are repeated iteratively until satisfactory results are obtained.

Parameter estimation is a major challenge in the iterative development of biochemical models.
Although parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
the complex dynamics of large biological systems and noisy, often incomplete experimental data sets pose a unique challenge.
Often optimization problems involving biological systems are non-linear and multi-modal i.e., typical models often have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}.
Non-linearity coupled with multi-modality generally renders local optimization techniques such as pattern search \cite{hooke1961direct}, Nelder-Mead simplex methods \cite{nelder1965simplex}, steepest descent methods or Levenberg-Marquardt \cite{more1978levenberg} incapable of reliably obtaining globally optimal solutions as these methods often terminate at local minimum.
Though deterministic global optimization techniques (for example algorithms based on branch and bound) can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global},
the absence of derivative information, the discontinuity of the objective functions, non-smooth regions or the lack of knowledge about the objective function hampers these techniques.

%Fan et al. recently showed that population based meta-heuristics along with decomposition based methods can also be used to identify gene circuit models from messenger RNA measurements \cite{fan2015parameter}.

Meta-heuristic stochastic optimization approaches like Genetic Algorithms (GAs) \cite{goldberg2006genetic}, Simulated Annealing (SA) \cite{kirkpatrick1983optimization},
Evolutionary Programming \cite{fogel2009artificial} and population based searches like Differential Evolution (DE) \cite{storn1997differential} have shown promise on nonlinear multi-modal problems \cite{sun2012parameter}.
These techniques do not make any assumptions about, nor do they require, \textit{a~priori} information about the structure of objective function.
Though they do not guarantee strong convergence, meta-heuristic approaches are often effective in finding globally optimal or near optimal solutions.
For example, Mendes et al. used simulated annealing to estimate rate constants for the irreversible inhibition of HIV proteinase \cite{mendes1998non},
while Modchang et al. used a genetic algorithm (GA) to estimate parameters for a model of G-protein-coupled receptors (GPCRs) activity \cite{modchang2008mathematical}. Parameter estimates obtained using GA helped qualitatively determine the effectiveness of two G-protein agonists, N6-cyclopentyladenosine (CPA) and 5'-N-ethylcarboxamidoadenosine (NECA).
Differential evolution (DE) have also been effective on various biological problems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}.
For example, Tashkova et al. compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis
and showed that DE was the most effective of the approaches tested \cite{tashkova2011parameter}.
Banga and co-workers have also successfully applied scatter-search methods to estimate parameters on non-linear biological processes \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter}.
Hybrid approaches that combine a meta-heuristic with a local optimization search,
where a near optimal solution estimated using a meta-heuristic is refined using a local search, have also become popular.
For example, Villaverde et al. developed the enhanced scatter search (eSS) method, which combined scatter search with local search methods,
for parameter estimation in a collection of systems biology models \cite{villaverde2015biopredyn}.
However, despite these successes, a major drawback of most meta-heuristics is the large number of function evaluations required to explore parameter space.
Performing a large number of potentially expensive function evaluations is not desirable (and perhaps not feasible) for many types of biochemical models.

%Typically as models grow in complexity, evaluation of the objective function becomes more computationally expensive.
%We fixed the number of function evaluations to 4000 for both the benchmark problems without pre-optimizing any algorithmic parameters.
%Lastly, we also compared the performance of DOPS against eSS on published benchmark problems \cite{villaverde2015biopredyn}.

In many high dimensional problems approaching an exact solution may not be necessary. Gutenkust et al. showed that a number of systems biology models are 'sloppy' \cite{gutenkunst2007universally}. Sloppy systems have specific parameter combinations that largely define the dynamics of the system. Large perturbations to the rest of the parameters does not greatly impact the system dynamics. Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}. Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions. DDS is a single solution based search where the solution is greedily updated. Because the neighborhood in DDS is the entire parameter space, DDS is a global optimization technique although it is greedy.

In this study, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel meta-heuristic that combines the global search capability of multi-swarm particle swarm optimization with the greedy refinement of dynamically dimensioned search.
The objective of DOPS is to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
DOPS uses multi-swarm particle swarm optimization to generate candidate solutions, which are then greedily updated using dynamically dimensioned search.
First, we tested the performance of DOPS on commonly used test functions for global optimization (Ackley and Rosenbrock) and published biochemical benchmark problems. Next, we used DOPS to identify a model of human coagulation cascade.
DOPS outperformed common meta-heuristic approaches like differential evolution, simulated annealing, and dynamically dimensioned search on the test functions, benchmark problems and the coagulation model. For example, DOPS recovered the nominal parameters for the benchmark problems using an order of magnitude less function evaluations than eSS in all cases. Taken together, these studies suggest DOPS is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.
